Predicting Diabetes Based on Various Factors Using Decision Tree Analysis

### 2023-12-14 manuscript/manuscript.tex
Title edit
Minor edits of intro and data

\documentclass[12pt]{article}
\usepackage{setspace}
\doublespacing

@@

\title{Predicting Diabetes Based on Various Factors Using Machine Learning Models}
\author{Ashley Merritt\\
    Department of Statistics\\
    University of Connecticut
    }

@@

    There are two types of diabetes that the majority of the patients have, type 1 or 2. The bodies of people that have type 1 diabetes
    completely do not make insulin, so they must take a daily insulin shot. Whereas in type 2 diabetes, these patients produce insulin but 
    their bodies do not use it properly (\cite{JDC2023Difference}). Therefore, it is very important that your body has a normal glucose level.
    In this paper, I will be working to determine the pre-existing factors that can be used in order to predict if a person may have diabetes. 
    Establishing the predictors is very important to me as many of my family members have a long history of suffering from this disease. 
    It is important to better predict this disease to save people from unnecessary suffering.

    Previously, the prediction of diabetes has been studied using machine learning models. Specifically in Sisodia's paper, it focused
    on the "Prediction of Diabetes using Classification Algorithms", they worked through the topic by producing results using support 
    vector machine, Naive Bayes classifier, and decision tree classifier (\cite{Sisodia2018Prediction}). In this paper, I will be 
    continuing the investigation with a different data set using decision trees.

    In this paper the specific research question that I will be focusing on is: how can we use machine learning models in order to identify 
    individuals that may have already or are at risk at developing diabetes? This will be researched using a data set that contains various 
    metrics on one's health that will be described in the next section of the paper.

@@

\section{Data}
\label{sec:data}
    In order to study the proposed research question on the topic of diabetes, I searched through many databases in order to find a set 
    that had all of the information I was looking for. After exploration, I found a data set entitled "Healthcare Diabetes Data set" on 
    the website Kaggle. The data set was originally sourced from the National Institute of Diabetes and Digestive and Kidney Diseases. The 
    data encompasses eight different predictors including pregnancies, glucose, blood pressure, skin thickness, insulin, BMI, diabetes 
    pedigree function, and age. In Table~\ref{tab:head_of_data} I will provide the first couple lines of dataset in order to provide
    more information on the dataset.

@@

  Another valuable piece of information for understanding our patients is seeing the range of the dataset. In Figure~\ref{fig:histogram}, we
  are able to see the spread of each predictor in our dataset. We can see that the majority of our patients are in the 20 to 30 year old range. 
  The bloodpressure predictor provides us with a somewhat normal distribution with the average being around 70 beats per minute. The BMI
  distribution is somewhat normal as well, with a slight right skew. Whereas the diabetes pedigree function is very skewed to the right. Glucose
  levels appear to be normal. Insulin levels appear to be very skewed to the right with the majority of values being 0. The outcome distribution
  is exactly as expected with only values of 0 or 1, from here on values of 0 will denote no detection of diabetes and 1 will detect diabetes. The
  number of pregnancies predictor also supplied a right skew but we know this to be valid as most people are not having 4 or more children. The 
  final predictor graphed was the skin thickness, this predictor displayed a right skew.

>>>>

\documentclass[12pt]{article}
\usepackage{setspace}
\singlespacing

@@

\title{Predicting Diabetes Based on Various Factors Using Decision Tree Analysis}
\author{Ashley Merritt\\
    Department of Statistics\\
    University of Connecticut
    }

@@

    There are two types of diabetes that the majority of patients have, type 1 or 2. The bodies of people that have type 1 diabetes
    completely do not make insulin, so they must take a daily insulin shot. Whereas individuals who have type 2 diabetes, produce insulin but 
    their bodies do not use it properly so they are treated either by exercise, taking on a diet or in some cases use of various medicines and/or insulin (\cite{JDC2023Difference}). 
    Therefore, it is very important that your body has a normal glucose level. In this paper, I will be working to determine the pre-existing
    factors that can be used in order to predict if a person may have diabetes. Establishing the predictors is very important to me as many of
    my family members have a long history of suffering from this disease. Therefore, this information could directly allow myself and my family members be
    proactive in terms of prevention and care. It is important to better predict this disease to save people from unnecessary suffering.

    Previously, the prediction of diabetes has been studied using other machine learning models. Specifically in Sisodia's paper, it focused
    on the "Prediction of Diabetes using Classification Algorithms", they worked through the topic by producing results using support 
    vector machine and Naive Bayes classifier. In Sisodia's paper they were able to get to an accuracy level of 76.3 percent, which we hope to exceed using
    a different machine learning model (\cite{Sisodia2018Prediction}). In this paper, I will be 
    continuing the investigation with a different data set using decision trees to better understand the predictors that help to determine if a patient
    may have diabetes.

    In this paper the specific research question that I will be focusing on is: how can we use machine learning models, specifically decision trees, in order to identify 
    individuals that may have already or are at risk at developing diabetes? This will be researched using a data set that contains various 
    metrics on one's health that will be described in the next section of the paper.

@@

\section{Data}
\label{sec:data}
    In order to study the proposed research question on the topic of diabetes, I searched through many databases in order to find a set 
    that had all of the information I was looking for. After exploration, I was able to find a data set entitled "Healthcare Diabetes Data set" on 
    the website Kaggle. The data set was originally sourced from the National Institute of Diabetes and Digestive and Kidney Diseases. The 
    data encompasses eight different predictors including pregnancies, glucose, blood pressure, skin thickness, insulin, BMI, diabetes 
    pedigree function, and age. In Table~\ref{tab:head_of_data} I will provide the first couple lines of dataset in order to provide
    more information.

@@

  Another valuable piece of information for understanding our patients is seeing the range of the dataset. In Figure~\ref{fig:histogram}, we
  are able to see the spread of each predictor in our dataset. We can see that the majority of our patients are in the 20 to 30 year old range and that the distribution is skewed to the right. 
  The bloodpressure predictor provides us with a slightly normal distribution with the average being around 70 beats per minute. The BMI
  distribution is somewhat normal as well, with a slight right skew. Whereas the diabetes pedigree function is very skewed to the right. Glucose
  levels appear to be normal. Insulin levels appear to be very skewed to the right with the majority of values being 0. The outcome distribution
  is exactly as expected with only values of 0 or 1, from here on values of 0 will denote no detection of diabetes and 1 will detect diabetes. The
  number of pregnancies predictor also supplied a right skew but we know this to be valid as most people are not having 4 or more children. The 
  final predictor graphed was the skin thickness, this predictor displayed a right skew.
